A Theory for Migration of Information Systems

### 2022-08-29 2022 Migration/articleMigration.tex
WIP on datasets

\section{Basics}
\label{sct:Basics}
	To describe migration, we first introduce the notion of dataset.
	A dataset describes a set of data, which is typically stored persistently in a database of some kind.
	Before defining datasets, we must first define the constituent notions of atom, concept, relation, rule, and role.
\subsection{Atoms}
	Atoms are values that have no internal structure, meant to represent data elements in a database.
	From a business perspective, atoms are used to represent concrete items of the world,
	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
	The set of all atoms is called $\atoms$.

\subsection{Concepts}
	Concepts are names we use to classify atoms in a meaningful way.
	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
	In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
	The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
	Specialization is needed to allow statements such as: ``A woman is a human'' or ``A human is a mammal''.
	The set of all concepts is called $\concepts$.

\subsection{Triples}
	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
	has \atom{074238991} as a telephone number.
	Triples are used to relate two atoms and a relation.
	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
	The practical meaning has no consequences in the formal world.
	The set of all triples is called $\triples$.

\subsection{Relations}
	Relations are used in information systems to store data.
	In this paper relations are represented by variables $r$, $s$, and $d$.

	Every relation $r$ has a set of pairs, which we call the population of $r$:
\begin{equation}
	\pairs{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\popR\}
\end{equation}
	Every relation has a name, a source concept, and a target concept.
	We write $\declare{nm}{A}{B}$ to denote a relation with name \id{nm}, source concept $A$, and target concept $B$.
	% To disassemble a relation in its name, source and target, there exist three functions:
% \[\begin{array}[3]{rcl}
	% \id{relname}&:&\rels\rightarrow\id{RelationIdentifier}\\
	% \id{src}&:&\rels\rightarrow\concepts\\
	% \id{tgt}&:&\rels\rightarrow\concepts
% \end{array}\]
	% Or, stated otherwise, the denotation $\declare{nm}{A}{B}$ stands for a relation with
	% \(\relname{\declare{nm}{A}{B}} = \id{nm}\),
	% \(\src{\declare{nm}{A}{B}}=A\), and
	% \(\tgt{\declare{nm}{A}{B}}=B\).
	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.

\subsection{Roles}
	A role is a name that identifies a group of users that get the same privileges in a data set.
	The set of all roles is called $\roles$.

\subsection{Constraints}
	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
	We call the set $\viol{\popR}{u}$ the \define{violation set} of constraint $u$ in population $\popR$.
	We call every element $\pair{a}{b}\in\viol{\popR}{u}$ a \define{violation} of constraint $u$ in $\popR$.
	Note that the signature of $u$ is the same for every population, whereas the violation set may be different for every $\popR$.

\subsection{Datasets}
	We can now define the notion of dataset.
	Let
	$\concepts$ be a set of concept symbols,
	$\rels$ be a set of relation symbols,
	$\roles$ be a set of roles,
	$\rules$ be a set of constraints, and
	$\atoms$ be a set of atoms.
\begin{definition}[dataset]
	\label{def:dataset}
	\item A dataset is a tuple $\la\concepts,\rels,\roles,\rules,\maintain,\atoms,\popR,\popC\ra$ with
	\[\begin{array}[3]{rcl}
		\id{maintain}&:&\roles\times\rules\\
		\id{popR}&:&\atoms\times\rels\times\atoms\\
		\id{popC}&:&\concepts\times\atoms\\
		\multicolumn{3}{l}{\triple{a}{r}{b}\in\popR\ \Rightarrow\ \pair{\src{r}}{a}\in\popC \wedge \pair{\tgt{r}}{b}\in\popC}\\
		\forall a\in\atoms\ \exists A\in\concepts&:&\pair{A}{a}\in\popC
	\end{array}\]
\end{definition}

>>>>

\section{Basics}
\label{sct:Basics}
	To define migration, let us first introduce the notion of dataset.
	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
	To abstract away from a particular database implementation,
	we represent data elements as triples.
	This makes our theory valid for any kind of database that triple stores can represent,
	such as SQL databases, graph databases and other no-SQL databases

	Before defining datasets, we must first define the constituent notions of atom, concept, relation, rule, and role.
\subsection{Atoms}
	Atoms are values that have no internal structure, meant to represent data elements in a database.
	From a business perspective, atoms are used to represent concrete items of the world,
	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
	The set of atoms is called $\atoms$.

\subsection{Concepts}
	Concepts are names we use to classify atoms in a meaningful way.
	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
	In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
	The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
	The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
	Specialization is needed to allow statements such as: ``A woman is a human'' or ``A human is a mammal''.
	The set of concepts is called $\concepts$.

\subsection{Triples}
	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
	has \atom{074238991} as a telephone number.
	Triples are used to relate two atoms and a relation.
	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
	The practical meaning has no consequences in the formal world.
	The set of triples is called $\triples$.

\subsection{Relations}
	Relations are used in datasets to store data.
	In this paper relations are represented by variables $r$, $s$, and $d$.

	Every relation $r$ has a set of pairs, which we call the population of $r$:
\begin{equation}
	\pop{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
\end{equation}
	Every relation has a name, a source concept, and a target concept.
	We write $\declare{nm}{A}{B}$ to denote a relation with name \id{nm}, source concept $A$, and target concept $B$.
	% To disassemble a relation in its name, source and target, there exist three functions:
% \[\begin{array}[3]{rcl}
	% \id{relname}&:&\rels\rightarrow\id{RelationIdentifier}\\
	% \id{src}&:&\rels\rightarrow\concepts\\
	% \id{tgt}&:&\rels\rightarrow\concepts
% \end{array}\]
	% Or, stated otherwise, the denotation $\declare{nm}{A}{B}$ stands for a relation with
	% \(\relname{\declare{nm}{A}{B}} = \id{nm}\),
	% \(\src{\declare{nm}{A}{B}}=A\), and
	% \(\tgt{\declare{nm}{A}{B}}=B\).

	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
	The set of relations is called $\rels$.

% \subsection{Roles}
% 	A role is a name that identifies a group of users that get the same privileges in a data set.
% 	The set of roles is called $\roles$.

\subsection{Constraints}
	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
	We call the set $\viol{\triples}{u}$ the \define{violation set} of constraint $u$ in population $\triples$.
	We call every element $\pair{a}{b}\in\viol{\triples}{u}$ a \define{violation} of constraint $u$ in $\triples$.
	Note that the signature of $u$ does not depend on the population, whereas the violation set of $u$ may be different for every $\triples$.
	The set of constraints is called $\rules$.

\subsection{Datasets}
	We can now define the notion of dataset.
	Let	$\concepts$ be a set of concept symbols,
	$\rels$ be a set of relations,
	$\atoms$ be a set of atoms,
	let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
	% let $\roles$ be a set of roles,
	and $\rules$ be a set of constraints.
\begin{definition}[dataset]
	\label{def:dataset}
	\item A dataset is a tuple $\la\concepts,\rels,\atoms,\inst,\triples,\rules\ra$ with
	\[\begin{array}[3]{rcl}
		% \id{maintain}&:&\roles\times\rules\\
		\forall\triple{a}{r}{b}\in\triples&:&a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
		\forall u\in\rules&:&\viol{\triples}{u}=\emptyset
	\end{array}\]
\end{definition}
